Forecasting biodiversity in breeding birds using best practices

### 2017-12-06 manuscript/reviewer_reply.Rmd

*In general, I thought this paper was interesting and well written. These results add to the increasing literature supporting the idea that ecological forecasting in time/space is difficult. I think the recommendations/best practice bullet points described by the authors are essential for anyone interested in forecasting this kind of data. All of my points below can be considered minor issues, which I think will improve the paper’s clarity.*

**Thank you.**

#### Introduction:
*Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much*

**This is a good point. We have added the word "terrestrial" to this line.**

@@

**RStan is cited in the last paragraph of the Methods.**

*Line 140: I’m curious why you didn’t consider spatial structure in the random effects here, using GMRF models or similar? One hand these models might be overkill if the habitat/environmental variables (which are spatially correlated) explain variability in richness, but the spatial correlation may pick up other sources of variability not accounted for by those variables.*

*Line 141-143: I think the paper thus far is pretty clear. I was hoping to see a few equations here – it helps me especially when components start to become hierarchical!*

**We have added the equations and Stan code as a supplemental file.**

@@

*Table 1: Just my personal preference, but I’d include the check marks and replace the ‘Xs’ with blank cells*

**We have made this change**

*Line 194-221: It seems for each approach you only considered the full models with all environmental covariates? I’m curious whether any of the covariates appeared to not improve explanatory power, and whether you considered models that dropped these uninformative predictors?*

**We chose not to drop uninformative predictors for two reasons. First, boosted trees are relatively robust to their inclusion ("A working guide to boosted regression trees", Elith et al. 2008), so it seemed unlikely to affect our results. Second, it would have added complexity to the model-fitting procedure, which already involved 500 rounds of cross-validation (one per Monte Carlo sample of the observer effect).**

@@

*Line 216: Cite randomForest / caret/ party R package – or whatever else you used?*

**The randomForest package was cited in the final methods paragraph. We've cited it where it is first mentioned as well.**

*Line 242: Again, I think including the equation here would be useful*

**We have added an equation and clarified the wording here.**

@@

*124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

**Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines [[XX]].**

*Fig. 2: Panels not labeled. No key for colors. Somewhat confusing.*

**Thank you for pointing this out.  As the Editor noted, we had inadvertently reversed the two panels. We have fixed this, and also labeled them as you suggested.**

### 2017-12-06 manuscript/draft.Rmd
misc updates/responses

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
[SDMs; @thomas2004; @thuiller2011; @urban2015]. In general, these models
describe individual species' occurrence patterns as functions of the
environment. Given forecasts for environmental conditions, these models can 
predict where each species will occur in the future. These species-level 
predictions are then combined ("stacked") to generate forecasts for species 
richness [e.g. @calabrese2014]. Alternatively, models that directly relate spatial 
patterns of species richness to environment conditions have been developed and 
generally perform equivalently to stacked SDMs [@algar2009; @distler2015]. 
This approach is sometimes referred to as "macroecological" modeling, because it 
models the larger-scale pattern (richness) directly [@distler2015].

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. For each forecasting
approach, we trained two versions of the corresponding model: one with
corrections for differences among observers, and one without (Figure 2). We
estimated the observer effects (and associated uncertainty about those effects)
with a linear mixed model, with observer as a random effect, built in the Stan
probabilistic programming language [@carpenter2017]. Because observer and site are
strongly related (observers tend to repeatedly sample the same site), site was
also included as a random effect to ensure that inferred deviations were
actually observer-related (as opposed to being related to the sites that a given
observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g.  variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
represented a different possible set of observer-level and site-level random
effect values across the full 32-year dataset.

>>>>

Previous efforts to predict future patterns of terrestrial species richness, and
diversity more generally, have focused primarily on building species
distributions models [SDMs; @thomas2004; @thuiller2011; @urban2015]. In general,
these models describe individual species' occurrence patterns as functions of
the environment. Given forecasts for environmental conditions, these models can
predict where each species will occur in the future. These species-level 
predictions are then combined ("stacked") to generate forecasts for species 
richness [e.g. @calabrese2014]. Alternatively, models that directly relate
spatial patterns of species richness to environment conditions have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. For each forecasting
approach, we trained two versions of the corresponding model: one with
corrections for differences among observers, and one without (Figure 2). We
estimated the observer effects (and associated uncertainty about those effects)
using a linear mixed model, with observer as a random effect, built in the Stan
probabilistic programming language [@carpenter2017]. Because observer and site
are strongly related (observers tend to repeatedly sample the same site),
site-level random effects were included to ensure that inferred deviations were
actually observer-related (as opposed to being related to the sites that a given
observer happened to see). The resulting model is described mathematically and
with code in Supplement S1. The model partitions the variance in observed
richness values into site-level variance, observer-level variance, and residual
variance (e.g. variation within a site from year to year).

Across our six modeling approaches (described below), we used estimates from the
observer model in three different ways.  First, the expected values for
site-level richness were used directly as our "average" baseline model (see
below). For the two models that made species-level predictions, the estimated
observer effects were included alongside the environmental variables as
predictors.  Finally, we trained the remaining models to predict
observer-corrected richness values (i.e. observed richness minus the observer
effect, or the number of species that would have been recorded by a "typical"
observer). Since the site-level and observer-level random effects are not known
precisely, we represented the range of possible values using 500 Monte Carlo
samples from the posterior distribution over these effects. Each downstream
model was then trained 500 times using different possible values for the random
effects.
